Robust Winner Determination and Simultaneous Elicitation of Scoring Rules and Preferences

### 2018-12-05 draft.tex
improved optimization section (plus small changes)

\DeclareMathOperator{\PMR}{PMR}
\DeclareMathOperator{\MR}{MR}
\DeclareMathOperator{\MMR}{MMR}

\title{Robust Winner Determination and Simultaneous Elicitation of Scoring Rules and Preferences}
\author{}
%\author{Paolo Viappiani} \affil{LIP6\\\email{paolo.viappiani@lip6.fr}}
%\author{Olivier Cailloux} 
%\author{Stefano Moretti}\affil{LAMSADE}

\begin{document}
\maketitle

%$\preflarge, \pref, \ppreflarge, \ppref$
% aggregate different opinions

%Minimax regret
%Given a full profile: use linear programming to maximize regret
%Given a partial profile: use techniques of Lu and Boutilier (2011).
Social choice deals with the problem of determining a consensus choice from the preferences of different agents (voters).
In the classical setting, the social choice rule is fixed beforehand; indeed many works analyze the properties of different rules (including axiomatic treatments) in order to justify the choice of specific social choice functions. 
Moreover, it is usually assumed that the preferences of the voters are completely known.

%We argue that it is possible that the committee wants to precise some preferences about the aggregation.
%We assume that obtaining information about them has a coast. 
%We  therefore aim at asking as few questions as possible.
In this draft we depart from the classic view by considering that both preferences and the social choice rule can be only partially specified.
We note that previous works have considered either partial preferences or a partially specified aggregation, but we do not know of any work considering both sources of uncertainty at the same time.
%We consider that the committee wants to express preferences for some voting rules
We provide a method for approximate winner determination and an incremental elicitation protocol based on minimax regret.

\section*{Background}

We assume that there is a set $A$ of $m$ alternatives and $n$ agents (voters); each agent is associated to a preference order.
%The set of preference orders is called profile and it is denoted by $v$.
The preferences of the agents are supposed to be linear orders (connected, transitive, asymmetric relations) involving the alternatives;
$\pref_i$ denotes the ``real'' preference relation of agent $i$. 
The set $(\pref_1,\ldots,\pref_n)$ is known in the social choice literature as the {\em preference  profile}.
The profile is equivalently represented by $\textbf{v}=(v_1,\ldots,v_n)$ where $v_i(j)$ denotes the rank (position) of alternative $j$ in the preference order $\pref_i$. 
With a little abuse of notation, we will use the term profile to refer to either $\textbf{v}$ or to the preference relations, depending on the context.
Let $V$ be the set of possible preference profiles (the cartesian product of $n$ linear orders).

A social choice function $f : V \rightarrow A$ associates a profile with one winner (or multiple winners in some cases).
Among the many possible social choice functions, we consider those rules that are based on a numerical score that is decomposable voter-wise.
%the score $s(a; v)$ of alternative $a$ is additive decomposable if it can be computed as $s(a; v) = \sum_{i} s(a; v_{i})$.
In particular positional scoring rules attach weights to positions;
an alternative obtains a score that depends on the rank obtained in each of the preference orders:

@@

\medskip
We also assume that the weights of the scoring rules are only partially specified.
We use $W$ to denote the set of feasible weight vector.
Without loss of generality, we assume $w_1=0$ and $w_m=0$.
The preferences of the chair are encoded with linear constraints, for instance one may state
that $w_2>0.5$.
A specific kind of preference may require the weights sequence to be convex, that means
that the difference between the weight of the first position and the weight of the second position is at least as much as the difference between the weights of the second and third positions, etc. 
\[ \forall i \in \{1,\ldots,n\} \;\; w_i - w_{i+1} \geq w_{i+1}-w_{i+2} \iff  w_i - 2 w_{i+1} + w_{i+2} \geq 0 \]
This is a constraint often used when aggregating rankings in sport competitions.

\section*{Related works}

@@

Lu and Boutilier \cite{Lu2011} assume that the preferences of the voters are only partially known (while the social choice function is known and fixed in advance; it is assumed to be decomposable) and 
 propose to use minimax regret to produce a robust approximation.
Each alternative is associated with a max regret value that measures how far from optimal it could be in the worst case given any completion of the partial profile.
The computation of max regret is facilitated by the fact the score is decomposable.
% We are given  $\ppref_i $,  a partial preference.
% The maximum regret is considered by assuming an adversary can choose the profile
% \begin{align*}
% \PMR(a,b; p) & = \max_{v \in C(p)} s(b; v) - s(a; v) \\
% \MR(a; p) & = \max_{b} \PMR(a,b) 
% \end{align*}
% 
% We then choose the minimax regret optimal alternative:
% \begin{align*}
% \MMR(p) & = \min \PMR(a,b) \\
% a^{*}(p) & \in \arg\min \PMR(a,b) 
% \end{align*}
%Max regret and minimax regret can be computed using independent completion of partial votes for non-decomposable scoring rules.

\section*{Incomplete profiles and partial information about the scoring rule}

{\bf Perhaps it could me made more general considering not only scoring rules, but all rules based on a score (as in Boutilier's paper)}

In this draft, we consider a setting where both the voters' preferences and the preferences of the chair about the voting rule are incomplete.
We write the score as $s(x; v,w)$ to underline the dependency of the score on the preference profile and on the weight vector. Let $s(x;v,w)$ denote the score of alternative $x$ in profile $\textbf{v}$ with weights $\textbf{w}$:
\[ s(x;\textbf{v},\textbf{w}) = \sum_{i=1}^{n} w_{\rank_i(x)} \]

%By using decomposition, we can compute the vote-wise regret...

@@

%We consider the pairwise max regret of alternative $x$  against alternative $y$ assuming that the weight vector belongs to $W$:

@@

\paragraph{Computation of minimax regret}

We will adapt the reasoning from \cite{Lu2011} combined with linear programming optimization.

Exploiting the decomposition of the score in terms of votes we cam rewrite $\PMR$ as follows:

\begin{align*}
\PMR(x,y; \pprofile, W) &= \max_{\w \in W} \max_{\profile \in C(\pprofile)} [ s(y; \profile,\w) - s(x; \profile,\w) ] = \\
&=  \max_{\w \in W} \sum_{j=1}^{n} \max_{v_j \in C(\succ_j^p)} [s(y; v_j,\w) - s(x; v_j,\w)]=\\
&=  \max_{\w \in W} \sum_{j=1}^{n} \max_{v_j \in C(\succ_j^p)} [w_{v_j(y)} - w_{v_j(x)}] \\
\end{align*}

{\bf \em This is just sketchy TODO: reason about the optimization and complete section}

In many cases the inner maximization can be made regardless of the outer max.
Part of the optimization can be made voter-wise: for each agent we extend the partial preference into a linear order that makes the gap between y and x as wide as possible.
However, in some cases the choice of how to complete the profiles may depend on the choice of weights, therefore it may be quite complicated...  (need to use integer variables???)
\medskip
 {\em Explain why convex sequences do not pose any problem}
In the case of convex sequences, the optimization can be done without integer variables.
Let $\hat{v}_k$ be the linear order extending $p_k$ according to the procedure described in \cite{Lu2011}.
\begin{align*}
\max_{\w \in W} \sum_{j=1}^{n}  [w_{\hat{v}_j(y)} - w_{\hat{v}_j(x)}]
\end{align*}
Therefore the pairwise maximum regret can be computed with a linear program.

\paragraph{Interactive Elicitation}
Starting from some initial partial knowledge, our goal is to learn both the scoring rule function and the agents' preferences.
While this is of course possible by doing one elicitation after another one, we propose an interleaved approach.
We adopt an interactive  protocol for simultaneously eliciting the preferences of the chair about the voting rule and the voters' preferences about the alternatives.
Indeed, it can be beneficial to interleave questions asked to the committee and questions asked to voters, depending on which is estimated to be more informative.

Answers given by the committee about the scoring rule refines our knowledge of the weights $w_1,\ldots,w_n$.
Answers given by one of the agents refine our knowledge about the agent's preferences

@@

Type of questions that we can ask to the chair:
bound queries, comparison queries.

@@

@@

>>>>

\DeclareMathOperator{\SCORE}{Score}
\DeclareMathOperator{\PMR}{PMR}
\DeclareMathOperator{\MR}{MR}
\DeclareMathOperator{\MMR}{MMR}

\title{Robust Winner Determination and Simultaneous Elicitation of Scoring Rules and Preferences}
%\author{}
%\author{Paolo Viappiani} \affil{LIP6\\\email{paolo.viappiani@lip6.fr}}
%\author{Olivier Cailloux} 
%\author{Stefano Moretti}%\affil{LAMSADE}
%\author{Beatrice Napolitano}\affil{LAMSADE}

\begin{document}
\maketitle
% aggregate different opinions

%Given a full profile: use linear programming to maximize regret
%Given a partial profile: use techniques of Lu and Boutilier (2011).
Social choice deals with the problem of determining a consensus choice from the preferences of different agents (voters).
In the classical setting, the social choice rule is fixed beforehand; indeed many works analyze the properties of different rules (including axiomatic treatments) in order to justify the choice of specific social choice functions. 
Moreover, it is usually assumed that the preferences of the voters are completely known.

%We argue that it is possible that the committee wants to precise some preferences about the aggregation.
In this draft we depart from the classic view by considering that both preferences and the social choice rule can be only partially specified.
We note that previous works have considered either partial preferences or a partially specified aggregation, but we do not know of any work considering both sources of uncertainty at the same time.
%We consider that the committee wants to express preferences for some voting rules
In this draft we provide a method for approximate winner determination and an incremental elicitation protocol based on minimax regret.

\section{Background}

{\bf We have to decide whether we want to focus only scoring rules, or on a larger class of rules based on a score.}

We assume that there is a set $A$ of $m$ alternatives and $n$ agents (voters); each agent is associated to a preference order.
%The set of preference orders is called profile and it is denoted by $v$.
The preferences of the agents are supposed to be linear orders (connected, transitive, asymmetric relations) involving the alternatives;
$\pref_i$ denotes the ``real'' preference relation of agent $i$. 
The set $(\pref_1,\ldots,\pref_n)$ is known in the social choice literature as the {\em preference  profile}.
The profile is equivalently represented by $\profile=(v_1,\ldots,v_n)$ where $v_i(j)$ denotes the rank (position) of alternative $j$ in the preference order $\pref_i$. 
With a little abuse of notation, we will use the term profile to refer to either $\profile$ or to the preference relations, depending on the context.
Let $V$ be the set of possible preference profiles (the cartesian product of $n$ linear orders).

A social choice function $f : V \rightarrow A$ associates a profile with one winner (or multiple winners in some cases).
Among the many possible social choice functions, we consider those rules that are based on a numerical score that is decomposable voter-wise.
%the score $s(a; v)$ of alternative $a$ is additive decomposable if it can be computed as $s(a; v) = \sum_{i} s(a; v_{i})$.
In particular {\em positional scoring rules} attach weights to positions; an alternative obtains a score that depends on the rank obtained in each of the preference orders:

@@

\medskip
We also assume that the weights of the scoring rules are only partially specified.
We use $W$ to denote the set of feasible weight vector.
Without loss of generality, we assume $w_1=0$ and $w_m=0$.
The preferences of the chair are encoded with linear constraints, for instance one may state
that $w_2>0.5$.
A specific kind of preference may require the weight sequence to be convex, that means that the difference between the weight of the first position and the weight of the second position is at least as much as the difference between the weights of the second and third positions, etc. 
\[ \forall i \in \{1,\ldots,n\} \;\; w_i - w_{i+1} \geq w_{i+1}-w_{i+2} \iff  w_i - 2 w_{i+1} + w_{i+2} \geq 0 \]
This is a constraint often used when aggregating rankings in sport competitions.

\section{Related works}

@@

Lu and Boutilier \cite{Lu2011} assume that the preferences of the voters are only partially known (while the social choice function is known and fixed in advance; it is assumed to be decomposable) and 
 propose to use minimax regret to produce a robust approximation.
Each alternative is associated with a max regret value that measures how far from optimal it could be in the worst case given any completion of the partial profile.
The computation of max regret is facilitated by the fact the score is decomposable.

% We are given  $\ppref_i $,  a partial preference.
% The maximum regret is considered by assuming an adversary can choose the profile
% \begin{align*}
% \PMR(a,b; p) & = \max_{v \in C(p)} s(b; v) - s(a; v) \\
% \MR(a; p) & = \max_{b} \PMR(a,b) 
% \end{align*}
% 
% We then choose the minimax regret optimal alternative:
% \begin{align*}
% \MMR(p) & = \min \PMR(a,b) \\
% a^{*}(p) & \in \arg\min \PMR(a,b) 
% \end{align*}
%Max regret and minimax regret can be computed using independent completion of partial votes for non-decomposable scoring rules.

\section{Incomplete profiles and partial information about the scoring rule}
In this draft, we consider a setting where both the voters' preferences and the preferences of the chair about the voting rule are incomplete.
Let $s(x;\profile,\w)$ denote the score of alternative $x$ in profile $\profile$ with weights $\w$.
Assuming positional scoring rules, the score of alternative $x$ is the sum of the points obtained 
\[ s(x;\profile,\w) = \sum_{i=1}^{n} w_{\rank_i(x)}. \]
We write the score as $s(x;\profile,\textbf{w})$ to underline the dependency of the score on the preference profile and on the weight vector.

The weights of a scoring rule can model different preferences for the committee; However in general it can be difficult to set the weights in an appropriate way.

@@

@@

\subsection{Computation of minimax regret}
{\bf \em TODO: reason about the optimization and complete section}

In order to compute pairwise maximum regret and therefore minimax regret, we adapt the reasoning from \cite{Lu2011} so that we decompose the PMR into the contributions that is associated to each agent.
The settings is however more challenging due to the presence of uncertainty in the weights.

Let $s(x; v_j,\w)=w_{v_j(x)}$ be the number of points $x$ obtains in the ranking $v_j$, that is the weight of the position $v_j(x)$.
Exploiting the decomposition of the score in terms of votes we cam rewrite $\PMR$ as follows:
\begin{align*}
\PMR(x,y; \pprofile, W) &= \max_{\w \in W} \max_{\profile \in C(\pprofile)} [ s(y; \profile,\w) - s(x; \profile,\w) ] = \\
&=  \max_{\w \in W} \sum_{j=1}^{n} \max_{v_j \in C(\succ_j^p)} [s(y; v_j,\w) - s(x; v_j,\w)]=\\
&=  \max_{\w \in W} \sum_{j=1}^{n} \max_{v_j \in C(\succ_j^p)} [w_{v_j(y)} - w_{v_j(x)}] \\
\end{align*}
%Computing PMR may be challenging since in principle the inner maximization can depend on the outer maximization.
Let $A^+$ be the set of agents for which we know that $y$ is preferred to $x$ (positive contribution to pairwise regret), $A^+$ 
be the set of agents for wich we  know that $x$ is preferred to $y$ (negative contribution) and $A^?$ the remaining case, where the preference between $x$ and $y$ is not known.
Consider how the adversary will complete the partial orders:
\begin{itemize}
 \item $j \in A^+$: If we know that agent $j$ prefers $y$ to $x$ then the adversary will complete the partial order of agent $j$ by placing as many as possible alternatives between $y$ and $x$.
 The way the partial order is completed in independent on the choice of the weights.

 \item $j \in A^-$: If we know that agent $j$ prefers $x$ to $y$ then the adversary will place as few as possible alternatives between $x$ and $y$.
 There is some dependecy between the weights and the completion of the partial order.
 This concerns the set of alternatives $U$ that are currently incomparable with $x$ and $y$.

 \item $j \in A^?$: If we do not know whether $x$ or $y$ is preferred by agent $j$ then the adversary will place $y$ before $x$ in the linear order and this case reduce to the first case; therefore also in this case the completion of the partial order can be done independently of the choice of $w$.
\end{itemize}

Let $\hat{v}_k$ be the linear order extending $p_k$ according to the above procedure (described in \cite{Lu2011}).

Then PMR can be written as follows:
\[ \PMR(x,y; \pprofile, W) = \max_{\w \in W} \Big \{ \max_{v_j \in C(\succ_j^p)} [w_{v_j(y)} - w_{v_j(x)}]  + \sum_{j \in A^+, j \in A^?} [w_{\hat{v}_j(y)} - w_{\hat{v}_j(x)}] \Big \} \]

\begin{itemize}
 \item Case of convex sequences:
  %{\em Explain why convex sequences do not pose any problem}
In the case of convex sequences, the optimization can be done without integer variables, since the case $A^-$ can be easily dealt with, and alternatives in $U$ can be indifferently be placed either to the top or to the bottom.
\begin{align*}
\max_{\w} & \sum_{j=1}^{n}  [w_{\hat{v}_j(y)} - w_{\hat{v}_j(x)}] \\
\text{s.t.} & \w \in W
\end{align*}
The space of feasible weights $W$ is encoded with linear constraints.
Therefore the pairwise maximum regret can be computed with a linear program.

 \item General case:
{\bf \em TODO: still have to find out exactly how to do it}
We introduce a set integer variables $I^i$, one for each position $i$.
 Let $U^j$ be, for the agents in $A^-$, the alternatives that, in the partial order $\succ^p_j$ of some agent $j \in A^-$, are incomparable with $x$ and $y$.
 \[ \sum_{l=1}^n I^j_l w_l \]
  Variable $I^j$ associated with the constraint $0 \leq I \leq |U^j|$.

  Use linearization techniques to handle the multiplication...
\end{itemize}

\subsection{Interactive Elicitation}
Starting from some initial partial knowledge, our goal is to learn both the scoring rule function and the agents' preferences.
While it is of course possible doing full elicitation of the weights and afterwards elicit the agents' preferences (or the other way around) we propose an interleaved approach.
In our interactive  protocol for simultaneously eliciting the preferences of the chair about the voting rule and the voters' preferences about the alternatives.
Indeed, it can be beneficial to interleave questions asked to the committee and questions asked to voters, depending on which is estimated to be more informative.

Answers given by the committee about the scoring rule refine our knowledge of the weights $w_1,\ldots,w_n$, while
answers given by one of the agents refine our knowledge about the agent's preferences.

@@

Type of questions that we can ask to the chair:
bound queries (Is this alternative among your top-k most preferred items?), comparison queries (do you prefer alternative x or alternative y?).

@@

\paragraph{Query methods}

@@

%We consider strategies of the following form
%{\em if} condition {\em then} ask comparison query {\em else} ask a committee query
%Define a score associated to each potential query that we may ask.

\begin{itemize}
\item Current solution strategy:
%Assume $x$ is the minimax regret optimal alternative $y$ the adversarial choice

\item Non-iterleaved regret method:

 \item One possibility is to consider the MMR {\em a posteriori}.
Assume that the different answers to a query induce the possible sets to be $(\pprofile_1,W_1)$ and $(\pprofile_2,W_2)$, then the score according to worst-case maximum regret is:
\[\SCORE(x)= \max_{i=1,2} \MMR(\pprofile_i,W_i) \]
In this case, the query with least value is chosen.
\end{itemize}

\section{Empirical Evaluation}

We evaluate the proposed method using the following simulated protocol.

\begin{itemize}
 \item Randomly generate the true preferences of the users (i.e. the linear orders) and the weights associated with the committee's preferences on the scoring rule
 \item For each of the elicitation strategy, simulate the elicitation by asking the queries selected by the strategy.

 \item Compare performance with respect to decrement in max regret, real loss
\end{itemize}

Test with different values of $m$ (number of alternatives) and $n$ (number of agents).

Test with different population sizes, different number of alternatives, etc.

Make one test assuming convex sequences of weights, and others without this assumption.

Comparison between our interleaved strategy and a strategy, still based on minimax regret  but that is not interleaved; also compare to some heuristic baselines.
